\documentclass{ltxdoc}
\usepackage{fontspec}
\usepackage[latin]{babel}
\usepackage{microtype}
\usepackage{hyperref}
\EnableCrossrefs
\CodelineIndex
\RecordChanges

\title{The \textsf{psalms} package\\Typesetting Latin psalms to Gregorian tones}
\author{Frank Barnes}
\date{2024-12-19}

\begin{document}
\maketitle

\begin{abstract}
The \textsf{psalms} package typesets Latin psalm texts with psalm tone
accents that follow the classic Gregorian mediant and termination formulas.
It combines Lua-based syllable detection with Lua\LaTeX\ for robust
accenting, and it ships with UTF-8 psalm texts for the 150 psalms plus
canticles used in the Divine Office.
\end{abstract}

\tableofcontents

\section{Introduction}

The package provides straightforward commands to select a psalm and tone
preset.  It expects UTF-8 encoded source texts and is intended for use with
Lua\LaTeX.  Example:
\begin{verbatim}
\documentclass{article}
\usepackage{fontspec}
\usepackage{psalms}
\begin{document}
\psalm{117}{8G}
\end{document}
\end{verbatim}

\section{Loading the package}

\begin{description}
  \item[Engine] Lua\LaTeX\ is required.  The package relies on Lua for
  syllabification and psalm tone logic.
  \item[Dependencies] \textsf{fontspec}, \textsf{babel} with Latin support,
  \textsf{xparse}, \textsf{enumitem}, and \textsf{lettrine}.
\end{description}

\section{Main command}

\DescribeMacro{\psalm}
\begin{quote}
\verb|\psalm{<number>}{<preset>}|
\end{quote}
The first argument is the psalm number.  The second chooses the tone preset
(for instance \verb|1|, \verb|2D|, \verb|7c|, or \verb|8G|).  Tone presets
are defined in the \texttt{psalmtones.lua} helper module.

\DescribeMacro{\psalmlegacy}
This legacy wrapper keeps compatibility with documents that previously used
\verb|\psalm[<preset>]{<number>}|.

\section{Package options}

\DescribeOption{orthographic}
Switches accent detection to orthographic mode, which looks for acute marks
(á, é, í, ó, ú, ý) or apostrophes.  Without marks, the positional fallback
(penultimate syllable) is used.

\DescribeOption{positional}
Explicitly sets positional mode (the default).

\DescribeOption{versenumbers}
Enables verse numbers globally.  Numbers are formatted with an
\textsf{enumitem} list that can be customised with
\verb|\SetPsalmVerseStyle|.

\DescribeOption{gloriapatri}
Appends the Gloria Patri to the end of every psalm.

\DescribeOption{dropcap}
Activates a decorative initial letter for the first verse using
\textsf{lettrine}.  Dropcap dimensions are controlled through the
\verb|\PsalmDropcap*| macros.

\DescribeOption{firstversegabc}
Renders only the first non-empty verse as a Gregorio snippet
\verb|\gabcsnippet{...}| generated from the selected tone preset.  When this
option is enabled, first-verse chant output takes precedence over
\verb|dropcap| for verse 1.  If the selected preset has no GABC definition or
marker mapping fails, the package falls back to the normal text-rendered verse.

\DescribeOption{debug}
Writes syllable diagnostics to the log file, which helps when tuning the
syllable splitter or tone presets.

\section{Run-time configuration}

The package exposes several helper macros:
\begin{description}
  \item[\verb|\SetPsalmAccentMode{<mode>}|] Switch accent logic at run time.
  \item[\verb|\SetPsalmVerseNumbers{true/false}|] Toggle verse numbers.
  \item[\verb|\SetPsalmGloriaPatri{true/false}|] Control the doxology.
  \item[\verb|\SetPsalmDropcap{true/false}|] Enable or disable dropcaps.
  \item[\verb|\SetPsalmFirstVerseGabc{true/false}|] Toggle first-verse chant
  rendering with \verb|\gabcsnippet| output.
  \item[\verb|\SuppressSyllabification|] Skip syllable recomputation for the
  next psalm, useful for fully premarked texts.
\end{description}

Typography helpers can be redefined as needed, e.g.\
\verb|\PsalmStyleAccent| to change highlighted syllables, or
\verb|\PsalmHalfDivider| to adjust the mediant divider.

\section{File layout}

Psalm texts live in the \texttt{tex/latex/psalms/psalms/} directory as UTF-8
\texttt{.txt} files.  Users can point the package to an alternate directory
by redefining \verb|\PsalmDir| or the \verb|\PsalmExt|.

The Lua helper \texttt{psalmtones.lua} resides in
\texttt{tex/lualatex/psalms/} and is loaded via \verb|\directlua|.

\section{First-verse chant markers}

The first-verse chant mode uses raw GABC cadence fragments in the Lua tone
table with marker tokens:
\begin{description}
  \item[\texttt{gr1}] Accented note group.  Accented sung syllables are mapped
  to these slots first.
  \item[\texttt{gr}] Extra note group for non-accent extension syllables.
  \item[Neutral notes] Fixed tone material that is emitted as-is and does not
  consume marker slots.
\end{description}

Markers are defined per cadence (flex, mediant, termination) for each preset
key (for example \texttt{1D}, \texttt{4E}, \texttt{8G}, \texttt{8G-star},
\texttt{peregrinus}).

\section{Example}

Normal text mode:
\begin{verbatim}
\documentclass{article}
\usepackage{fontspec}
\usepackage{psalms}
\begin{document}
\psalm{117}{8G}
\end{document}
\end{verbatim}

First-verse chant mode with the same preset:
\begin{verbatim}
\documentclass{article}
\usepackage{fontspec}
\usepackage[firstversegabc]{psalms}
\begin{document}
\psalm{117}{8G}
\end{document}
\end{verbatim}

\section{Building the documentation}

Run \verb|lualatex psalms-doc.tex| twice to resolve cross references.  The
package documentation depends on \textsf{ltxdoc} and the same runtime
requirements as the package itself.

\StopEventually{}
\PrintChanges
\PrintIndex

\end{document}
